Dual host-parasite transcriptomes of apicomplexan Eimeria falciformis and its natural mouse host

### 2016-07-20 figuresANDmanuscript/manuscriptBMCformat/manuscript_rnaseq.tex

%%% Load packages
%\usepackage{amsthm,amsmath}
%\RequirePackage{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
%\RequirePackage{hyperref}
%\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
%\usepackage{siunitx}

@@

\usepackage[colorlinks]{hyperref} % To color blocks of text 
%\usepackage[dvipsnames]{xcolor} % More color options (see Wikibooks)  
\usepackage{changepage}
\usepackage{booktabs} % For prettier tables
\usepackage{float} % To control positions of e.g. tables or figures
\usepackage[section]{placeins}
\usepackage{graphicx}
\graphicspath{ {../} }
%\def\includegraphic{}
%\def\includegraphics{}

@@

\author[
   addressref={aff1},
]{\inits{TK}\fnm{Totta} \snm{Kasemo}}
\author[
   addressref={aff1},
]{\inits{SS}\fnm{Simone} \snm{Spork}}
\author[
   addressref={aff2},
]{\inits{CD}\fnm{Christoph} \snm{Dieterich}}
\author[
   addressref={aff1},
]{\inits{RL}\fnm{Richard} \snm{Lucius}}
\author[
   addressref={aff1, aff3},                   % id's of addresses, e.g. {aff1,aff2}
   corref={aff1},                       % id of corresponding address, if any
   noteref={n1},                        % id's of article notes, if any
   email={emanuel.heitlinger@hu-berlin.de}   % email address
]{\inits{EH}\fnm{Emanuel} \snm{Heitlinger}}

@@

\address[id=aff1]{%                           % unique id
  \orgname{Department of Biology, Humboldt-Universitat zu Berlin}, % university, etc
  \street{Philippstr. 13, Haus 14},                     %
  %\postcode{}                                % post or zip code
  \city{Berlin},                              % city
  \cny{Germany}                                    % country
}
\address[id=aff2]{%  
  \orgname{Max Planck Institute for Biology of Ageing},
  \street{Joseph-Stelzmann-Strasse 9B},
  \postcode{50931}
  \city{Cologne},
  \cny{Germany}
}
\address[id=aff3]{%  
  \orgname{Department Reproduction Biology and Evolutionary Ecology, Leibniz Institute of Zoo and Wildlife Research},
  \street{Alfred-Kowalke-Str. 17},
  \postcode{10315}
  \city{Berlin},
  \cny{Germany}
}

@@

%\begin{artnotes}
%\note{Sample of title note}     % note to the article
%\note[id=n1]{Equal contributor} % note, connected to author
%\end{artnotes}

@@

\begin{keyword}
\kwd{Parasite, apicomplexa, RNA-seq, transcriptome, life-cycle, interaction}
\kwd{article}
\kwd{author}
\end{keyword}

@@

\end{abstractbox}
%
\end{fmbox}% uncomment this for twcolumn layout

\end{frontmatter}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% 	NUMBER OF GENES DIFFERENT IN DIFFERENT COMPARISONS	
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
\section{mRNA abundance differences between different experimental groups}
\setlength{\tabcolsep}{8pt}
\begin{table}[H]
\small
\begin{center}
\caption{mRNA abundance differences between different experimental groups.}
%\label{tab:table3}
\begin{tabular}{*3l}    \toprule
\textit{Day post infection} & \textit{Ef} genes different & Mouse genes different \\ 
	\textit{comparisons} 	    & (FDR$\leq$0.01) &  (FDR$\leq$0.01/0.05) \\ \midrule
	NMRI 0 vs NMRI 3		& NA   & 274 \\
	NMRI 0 vs NMRI 5		& NA   & 1736 \\
	NMRI 0 vs NMRI 7		& NA   & 2802 \\
	NMRI 3 vs NMRI 5     		& 111  & 1 \\
	NMRI 3 vs NMRI 7  		& 1385 & 1407 \\ 
	NMRI 5 vs NMRI 7  		& 1895 & 873 \\ 
	C57BL/6 0 vs C57BL/6 5		& NA	& 914 \\
	Rag1-/- 0 vs Rag1-/- 5		& NA	& 45 \\ \midrule
\textit{Day post infection,} & 		 & 		 \\ 
\textit{parasite relevant comparisons} 	    & 		&  \\ \midrule
	Oocysts vs NMRI 3  	& 3310 & NA \\  
	Oocysts vs NMRI 5	& 3605 & NA \\ 
	Oocysts vs NMRI 7	& 3085 & NA \\ 
	Oocysts vs sporozoites  & 3421 & NA \\
	Sporozoites vs NMRI 3 	& 1663 & NA \\
	Sporozoites vs NMRI 5 	& 1605 & NA \\
	Sporozoites vs NMRI 7 	& 2473 & NA \\ 
\textit{First and second infection} & 		 & 	 \\ 
\textit{comparisons} 	    & 		& 	\\ \midrule
	NMRI 3 1st vs NMRI 3 2nd  	& 0  & 5 \\
	NMRI 5 1st vs NMRI 5 2nd  	& 0  & 1 \\
	NMRI 7 1st vs NMRI 7 2nd  	& 0  & 902 \\
	C57BL/6 1st vs C57BL/6 2nd (day 5) & 0 &  mouse \\
	Rag1-/- 1st vs Rag1-/- 2nd (day 5) & 0 & mouse  \\ \midrule
\textit{Mouse strain comparisons} & 		 & 	 \\ \midrule
	NMRI vs C57BL/6 		& 22 	& NA* \\		
	NMRI vs Rag1-/-     		& 0 & NA* 	\\
	C57BL/6 vs Rag1-/-  		& 0   & 356 	\\ \midrule
\bottomrule
	\hline
\end{tabular}
\end{center}
\end{table}

@@

\subsubsection{Data normalisation and sample exclusions}
Count data was normalised in R, using the upperquartile method as implemented in edgeR (version 3.14.0). Briefly, normalisation assumes a negative binomial distribution of data and dispersed variance for genes with higher mean abundance. The normalisation shrinks the dispersion of variance towards a Poisson (linear) relationship between mean and variance. We evaluated the density distribution of our data and two samples contradicted the assumed negative binomial distribution and were excluded from further analysis: NMRI\_1st\_3dpi\_rep1 and NMRI\_2nd\_5dpi\_rep1 (SI ...).

\subsection{Differentially abundant mRNAs used for hierarchical clustering}
A selection of differently abundant mRNAs are used for hierarchical clustering of \textit{E. falciformis} life cycle relevant genes. In each comparison (see Table 3), the 500 differentially abundant mRNAs with lowest FDR are selected. In the next step, the 500 mRNAs from each comparison (or less) are joined (4935 genes) and only unique genes are selected (1618 genes). The 22 genes in the NMRI vs C57BL/6 comparison are not included in the \textit{E. falciformis} life cycle analysis.

All parasite samples were included in heatmap of genes detected as differentially abundant in timepoint comparisons.

On scale bar in heatmaps, 0 is mean mRNA abundance for each gene (row). Up (green) and down-regulation (brown) numbers denote number of standard deviations from row mean. Hierarchical clustering was performed using with Euclidean distances, method 'complete' (R package base).

\setlength{\tabcolsep}{10pt}
\begin{table}[H]
\small
\begin{center}
\caption{Differentially abundant \textit{E. falciformis} mRNAs used for hierarchical clustering.}
\begin{tabular}{*2l}    \toprule
	\textit{Data description} & Number of genes \\ \midrule
	Sum of 1st infection NMRI sample differences	& 4935  \\ 
	(including oocysts and sporozoites)	\\	
	Used in hierarchical clustering (heatmap)  	& 1618 \\ 	\bottomrule	
\hline
\end{tabular}
\end{center}
%\end{adjustwidth}
\end{table}

>>>>

%%% Load packages
%\usepackage{amsthm,amsmath}
%\RequirePackage{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
%\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
%\usepackage{siunitx}
\usepackage[colorlinks]{hyperref} % To color blocks of text 
%\usepackage[dvipsnames]{xcolor} % More color options (see Wikibooks)  
\usepackage{changepage}
\usepackage{booktabs} % For prettier tables
\usepackage{float} % To control positions of e.g. tables or figures
%\usepackage[section]{placeins}
\usepackage{graphicx}
\graphicspath{ {../} }

@@

%\def\includegraphic{}
%\def\includegraphics{}

@@

\author[
%	addressref={aff1},
%	noteref={n1},
]{\inits{TK}\fnm{Totta} \snm{Kasemo}}
\author[
%	addressref={aff1},
]{\inits{SS}\fnm{Simone} \snm{Spork}}
\author[
%	addressref={aff2},
%	noteref={n1},
]{\inits{CD}\fnm{Christoph} \snm{Dieterich}}
\author[
%	addressref={aff1},
]{\inits{RL}\fnm{Richard} \snm{Lucius}}
\author[
%	addressref={aff1, aff3},                   % id's of addresses, e.g. {aff1,aff2}
	corref={aff1, aff3},                       % id of corresponding address, if any
%	noteref={n1},                        % id's of article notes, if any
	email={emanuel.heitlinger@hu-berlin.de}   % email address
]{\inits{EH}\fnm{Emanuel} \snm{Heitlinger}}

@@

\address[id=aff1]{%                           % unique id
	\orgname{Department of Biology, Humboldt-Universitat zu Berlin}, % university, etc
	\street{Philippstr. 13, Haus 14},                     %
	\postcode{10115}                                % post or zip code
	\city{Berlin},                              % city
	\cny{Germany}                                    % country
}
\address[id=aff2]{%  
	\orgname{Max Planck Institute for Biology of Ageing},
	\street{Joseph-Stelzmann-Strasse 9B},
	\postcode{50931}
	\city{Cologne},
	\cny{Germany}
}
\address[id=aff3]{%  
	\orgname{Department Reproduction Biology and Evolutionary Ecology, Leibniz Institute of Zoo and Wildlife Research},
	\street{Alfred-Kowalke-Str. 17},
	\postcode{10315}
	\city{Berlin},
	\cny{Germany}
}

@@

\begin{artnotes}
%\note{Sample of title note}     % note to the article
%\note[id=n1]{Equal contributor} % note, connected to author
%\note[id=n2]
\end{artnotes}

@@

\begin{keyword}
\kwd{Parasite, apicomplexa, RNA-seq, transcriptome, life-cycle, interaction}
%\kwd{article}
%\kwd{author}
\end{keyword}

@@

\end{abstractbox}
\end{fmbox}% uncomment this for twcolumn layout
\end{frontmatter}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% 	NUMBER OF GENES DIFFERENT IN DIFFERENT COMPARISONS	
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
\section{mRNA abundance differences between different experimental groups}
\setlength{\tabcolsep}{8pt}
\begin{table}[H]
\small
\begin{center}
\caption{mRNA abundance differences between different experimental groups.}
%\label{tab:table3}
\begin{tabular}{*3l}    \toprule
\textit{Day post infection} & \textit{Ef} genes different & Mouse genes different \\ 
	\textit{comparisons} 	    & (FDR$\leq$0.01) &  (FDR$\leq$0.01/0.05) \\ \midrule
	NMRI 0 vs NMRI 3		& NA   & 274 \\
	NMRI 0 vs NMRI 5		& NA   & 1736 \\
	NMRI 0 vs NMRI 7		& NA   & 2802 \\
	NMRI 3 vs NMRI 5     		& 111  & 1 \\
	NMRI 3 vs NMRI 7  		& 1385 & 1407 \\ 
	NMRI 5 vs NMRI 7  		& 1895 & 873 \\ 
	C57BL/6 0 vs C57BL/6 5		& NA	& 914 \\
	Rag1-/- 0 vs Rag1-/- 5		& NA	& 45 \\ \midrule
\textit{Day post infection,} & 		 & 		 \\ 
\textit{parasite relevant comparisons} 	    & 		&  \\ \midrule
	Oocysts vs NMRI 3  	& 3310 & NA \\  
	Oocysts vs NMRI 5	& 3605 & NA \\ 
	Oocysts vs NMRI 7	& 3085 & NA \\ 
	Oocysts vs sporozoites  & 3421 & NA \\
	Sporozoites vs NMRI 3 	& 1663 & NA \\
	Sporozoites vs NMRI 5 	& 1605 & NA \\
	Sporozoites vs NMRI 7 	& 2473 & NA \\ \midrule
\textit{First and second infection} & 		 & 	 \\ 
\textit{comparisons} 	    & 		& 	\\ \midrule
	NMRI 3 1st vs NMRI 3 2nd  	& 0  & 5 \\
	NMRI 5 1st vs NMRI 5 2nd  	& 0  & 1 \\
	NMRI 7 1st vs NMRI 7 2nd  	& 0  & 902 \\
	C57BL/6 1st vs C57BL/6 2nd (day 5) & 0 &  mouse \\
	Rag1-/- 1st vs Rag1-/- 2nd (day 5) & 0 & mouse  \\ \midrule
\textit{Mouse strain comparisons} & 		 & 	 \\ \midrule
	NMRI vs C57BL/6 		& 22 	& NA* \\		
	NMRI vs Rag1-/-     		& 0 & NA* 	\\
	C57BL/6 vs Rag1-/-  		& 0   & 356 	\\ \midrule
\bottomrule
	\hline
\end{tabular}
\end{center}
\end{table}

@@

\subsection{Differential mRNA abundance, data normalisation and sample exclusions}
Complete scripts are available at git@github.com:derele/Ef\_RNAseq.git. Count data was analysed in R, using the upperquartile normalisation method as implemented in edgeR (version 3.14.0). Briefly, genes with zero coverage in all samples (libraries) are removed and normalisation factors are calculated for the 75\% quantile for each library. This  normalisation assumes a negative binomial distribution of read densities. Two samples contradicted the assumption (parasite data) and both mouse and parasite data from these samples were excluded from further analysis: NMRI\_1st\_3dpi\_rep1 and NMRI\_2nd\_5dpi\_rep1 (SI ...). We used a the edgeR pipleine which estimates dispersion values for each gene (provided that a design matrix is defined). The method fits data for each gene to a negative binomial distribution using a generalized linear model (GLM) with the function glmFit() and glmLRT().

........... include outcome of MDS analysis here?

\subsection{Differentially abundant mRNAs used for hierarchical clustering}
A selection of differently abundant mRNAs are used for hierarchical clustering of \textit{E. falciformis} life cycle relevant genes. In each comparison (see Table 3), the union of the 500 differentially abundant mRNAs with lowest FDR are selected. In the next step, the 500 mRNAs from each comparison (or less) are joined. For \textit{E. falciformis} time p.i. comparison, this resulted in 4935 genes of which only the 1618 unique genes were selected. The 22 genes in the NMRI vs C57BL/6 comparison are not included in the \textit{E. falciformis} life cycle analysis. The same comparisons for mouse yielded 8052 genes in total and 1313 unique ones. In heatmaps, all samples, i.e. also samples which themselves did not have any significantly different mRNAs according to our selection, were included in hierarchical clustering. Scale bar in heatmaps show 0 as mean mRNA abundance for each gene (row). Up (green) and down-regulation (brown) denote number of standard deviations from 0, i.e., row mean. Hierarchical clustering was performed using with Euclidean distances, using complete linkage ('complete', R package base).

\setlength{\tabcolsep}{10pt}
\begin{table}[H]
\small
\begin{center}
\caption{Genes used for hierarchical clustering of mRNAs differently abundant depending on time p.i..}
\begin{tabular}{*3l}    \toprule
	\textit{Data description} & \emph{E. falciformis} genes & Mouse genes	\\ \midrule
	Sum of 1st infection NMRI sample differences	& 4935 & 8052 \\ 
	(including oocysts and sporozoites if appl.)	\\	
	Used in hierarchical clustering (heatmap)  	& 1618 & 1313 \\ 	\bottomrule	
\hline
\end{tabular}
\end{center}
%\end{adjustwidth}
\end{table}
